---
title: "re_LULC"
author: "Flueckiger_Rubens"
date: "2025-11-11"
output: 
  html_document:
    toc: TRUE
---

# Packages

```{r}
library(tidyr)
library(dplyr)
library(parsnip)
library(workflows)
library(tune)
library(dials)
library(rsample)
```

# Load df's
Load dataset from website and safe it in folder

```{r}
if(FALSE){
  # see here how to directly load train and test data from Zenodo
  download.file(
    url = "https://zenodo.org/records/8298491/files/test_data.rds?download=1",     
    destfile = file.path("../data_raw/", "test_data.rds"))
  download.file(
    url = "https://zenodo.org/records/8298491/files/training_data.rds?download=1", 
    destfile = file.path("../data_raw/", "training_data.rds"))
}

```

Note: If it does not work files can be downloaded here: <https://zenodo.org/records/8298491>

Load data set from folder

```{r}

test_data_Zenodo <- readr::read_rds(
  file.path("../data_raw/", "test_data.rds"))
training_data_Zenodo <- readr::read_rds(
  file.path("../data_raw/", "training_data.rds"))

# ensure training and test data contain same predictor/feature variables
setdiff(names(training_data_Zenodo), 
        names(test_data_Zenodo)) # pixelID, LC1, lat, lon
setdiff(names(test_data_Zenodo),     
        names(training_data_Zenodo))

train <- training_data_Zenodo |> 
  select(-c(pixelID, lat, lon)) # remove ID, lat, lon
test  <- test_data_Zenodo                                           
# NOTE: for the competition, column 'LC1' was removed from the 
#       test data
```

set seed for reproduceability

```{r}
set.seed(1)
```

# Imporvements

## Ideas

1.  KNN
2.  Hyperparameters
3.  # Models (by: Crossvalidation, Hypercube adjustment thingy)
4.  LOESS for selecting best model

## Create model

```{r}
model_settings <- parsnip::boost_tree(
  engine = "xgboost",
  trees = 50,
  min_n = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_mode("classification")

```

Workflow

```{r}
xgb_workflow <- workflows::workflow() |>
  add_formula(as.factor(LC1) ~ .) |>
  add_model(model_settings)

print(xgb_workflow)
```

## Hypercube

```{r}
hp_settings <- dials::grid_space_filling(
  tune::extract_parameter_set_dials(xgb_workflow),
  size = 5,
  type = "latin_hypercube"
)

print(hp_settings)
```

## Parameter tuning

note: Duration is long (I needed ~ 7h). So... store result as RDS format in folder
note: seed for reproduceability needed

```{r}
if(FALSE){
  xgb_results <- tune::tune_grid(
    xgb_workflow,
    resamples = vfold_cv(data = train, v = 5),
    grid = hp_settings,
    control = tune::control_grid(
      save_pred = TRUE, # Save pred. statistics
      verbose = TRUE # Show progress while running
    )
  )
  
  # Save final data set
  saveRDS(object = xgb_results, file = "../model_data/xgb_result")
}
```

## Select best model

Load result from folder, select best according to LOESS, finalize workflow

```{r}
xgb_results <- readr::read_rds(
  file.path("../model_data/", "xgb_result")
)

xgb_best <- tune::select_best(
  xgb_results,
  metric = "roc_auc"
)

xgb_best_hp <- tune::finalize_workflow(
  xgb_workflow,
  xgb_best
)

print(xgb_best_hp)
```

## Fit

Again save output since running takes for ever

```{r}
if(FALSE){
  # Fit best model
  xgb_best_model <- fit(xgb_best_hp, train)
  
  # Save model
  saveRDS(xgb_best_model, file = "../model_data/xgb_best_model")
}
```

Load data set

```{r}
xgb_best_model <- readr::read_rds(
  file.path("../model_data/", "xgb_best_model")
)
```

predict result

```{r}
test_results <- predict(xgb_best_model, test)
```

Save the results for the leaderboard

```{r}
readr::write_csv(
  x = dplyr::rename(test_results,
                    lulc_class = .pred_class),
  file = "~/Desktop/VanEend_results.csv")
```