---
title: "Spatial Upscalling"
author: "Flueckiger_Rubens"
date: "2025-11-24"
output: 
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
      smooth_scroll: FALSE
---

# 1 Introduction

Cross validation (CV) is an important feature in machine learning to optimise target prediction. However, a lack of sufficiently spatially covered observation data yields a big problem for accurately predicting a target variable in a observation-sparse region. In this report exercise, I will compare three different methods of splitting training data in n different folds to tackle the previous mentioned problem. First, CV-folds will be generated by randomly splitting training data in 5 folds. Secondly, I will split the training data into 5 regions according to the observations geographical location. These splits will be used to perform a 5-fold CV. Finally, I will split the training data into 5 classes according to the predictors temperature and precipitation value to perform a 5-fold CV. For each CV $R^2$ and Root mean square error ($RMSE$) value will be returned to compare the results with each other.

# 2 Literature analysis

Given the paper of Ludwig et al. ([2023](https://doi.org/10.1111/geb.13635)) we will first answer the following questions; What is the difference between a *random* and a *spatial* CV, and could one use a different metric other than the euclidean distance metric used in the paper to group observations into different classes.

## 2.1 CV differences

A random cross-validation method splits data independent of their location into n training folds; the resulting folds only match in their distribution of predictor variables. This approach lead to variables that can well reproduce clustered training data but might not be well-suited for spatial predictions due to an insufficiently covered training data to predict remote location settings. Spatial cross-validation on the other hand distributes data into different folds according to their geographical location. This approach might mimic a real-world scenario a bit better as it will favour variables that lead to a high ability to make predictions for new areas.

## 2.2 Distance operator

The goal of using a euclidean distance operator is to use a metric, in this case it is the location of the observation given in longitude and latitude, to group an observation into a certain class. The underlying assumption is thereby that observations that are close to each other will encounter the same climatological conditions, thereby lowering the prediction error. However, as we want to predict a certain target variable with given observable predictors, such as temperature, solar irradiance, precipitation or pressure, one could also use a subset of these predictors to group observations into different classes directly. This approach will yield the same effect as the distance operator approach since observations under similar climatological conditions (in our case we limited the climatological conditions inside the selected subset of the predictors), will be grouped into the same class.

# 3 Packages and data

In a first step, all necessary packages to perform the analysis will be loaded.

```{r packages, message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(ggplot2)
library(ranger)
library(RColorBrewer)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(skimr)
library(tidyverse)
```

In a next step we can load the data set. Since the data set will be loaded from the internet, I will store it additionally inside the *data_raw* folder such that the file can be reproduced without internet access. To ensure that the file loading process will not be executed again it will be enclosed within a `if(FALSE){}` statement.

```{r dataframe, message=FALSE}
if(FALSE){
  df <- readr::read_csv(
    "https://raw.githubusercontent.com/geco-bern/leafnp_data/main/data/leafnp_tian_et_al.csv")
  
  # Safe df for offline use
  write.csv(
    df,
    file = "../data_raw/leafnp_data", # Location
    row.names = FALSE # Don't save indices
  )
}

df <- readr::read_csv("../data_raw/leafnp_data")
```

## 3.1 Data preprocessing

Now we will create a subset of the original data, which only contains the necessary predictors, which are:

-   `leafN`: leaf nitrogen content, in mass-based concentration units [$gNgDM^{-1}$]
-   `lon`: longitude in decimal degrees east
-   `lat`: latitude in decimal degrees north
-   `elv`: Elevation above sea level [$m$]
-   `mat`: mean annual temperature [$°C$]
-   `map`: mean annual precipitation [$mmyr^{-1}$]
-   `ndep`: atmospheric nitrogen deposition [$gmyr^{-1}$]
-   `mai`: mean annual daily irradiance [$\mu mol m^{-2} s^{-1}$]
-   `Species`: species name of the plant on which leaf N was measured

```{r df preprocessing}
common_species <- df |>
  group_by(Species) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  slice(1:50) |>
  pull(Species)

dfs <- df |>
  dplyr::select(leafN, lon, lat, elv, mat, map, ndep, mai, Species) |>
  filter(Species %in% common_species)

# quick overview of data
skimr::skim(dfs)
```

# 4 Random CV

Given the data, we will now perform a random cross-validation. To ensure reproducibility, especially for the interpretation later on, I will set a seed.

```{r seed}
set.seed(102)
```

I will use the `caret` package to perform the random CV. Thereby we first need to specify the model given the target variable (leaf nitrogen) and predictors (elevation, temperature, precipitation, atmospheric nitrogen content, solar irradiance and leaf nitrogen measured species) and convert the string into a `caret`-readable format using the base R `formula()` function.

```{r model preprocessing}
# Select predictors
predictors <- names(dfs[4:ncol(dfs)])

# Create string with all variables
fun_pred <- paste(
  c("leafN ~", paste(predictors, collapse = " + ")),
  collapse = " "
)

# Convert into a caret-readable string
fun_pred <- formula(fun_pred)
```

With the `carret::train()` function we can now specify the model to be performed. I will use the model function created above combined with the preprocessed data from Chapter [3.1 Data preprocessing]. We will choose method as ranger to perform a 5-fold CV. Inside the `tuneGrid()` function we will specify the model parameters. I will set `mtry` to be equal to 3, `min.node.size` to be 12 and the remaining parameters as their default in `ranger` (which is only `splitrule`; default: 'variance'), as required by the task. To return the values of $R^2$ and $RMSE$, simply using `$resample` will return the wished metrics for each fold.

```{r random cv}
mod_cv <- train(
  fun_pred, # Model function
  data = dfs, # training set
  method = "ranger", # Random Forest model
  trControl = trainControl(
    method = "cv", # Cross-Validation method
    number = 5 # 5-Fold CV
  ),
  tuneGrid = data.frame(
    .mtry = 3, # 
    .min.node.size = 12,
    .splitrule = "variance" # Default
  )
)

# Return metrics
stat <- mod_cv$resample[c("Rsquared", "RMSE")]
stat <- stat |> rename( # Rename row names
  "Rsq_ran" = "Rsquared",
  "RMSE_ran" = "RMSE"
) |> round(3) |> mutate( # Round observations
  "Fold" = mod_cv$resample["Resample"]# Fold number
) 

stat <- stat[c("Fold", "Rsq_ran", "RMSE_ran")] # Reorder columns

knitr::kable(
  stat, 
  align = "lcc",
  caption = "Table 2: R^2 and RMSE values from a random 5-fold CV"
)
```

# 5 Spatial CV

In a next step we will perform a spatial CV using geographical location as the determining factor for splitting training data in 5 different folds.

## 5.1 Observation distribution

First we will analyse the spatial distribution of our observations.

```{r data distribution, fig.align='center', fig.cap="Figure 1: Spatial distribution of every observation of the data set"}
# get coast outline
coast <- rnaturalearth::ne_coastline(scale = 110, returnclass = "sf")

ggplot() +
  
  #Plot coastline
  geom_sf(data = coast, colour = "black", size = 0.2) +

  # set extent in longitude and latitude
  # to draw map strictly bounded by the specified extent
  coord_sf(ylim = c(-60, 80), expand = FALSE) + 

  # plot points on map
  geom_point(data = dfs, aes(x = lon, y = lat), color = "red", size = 0.2) +
  
  # Lable adjustments
  labs(
    x = "", y = "",
    title = "Spatial distribution of observations used in the dataset"
  ) +
  
  theme_bw() +
  theme(legend.position = "bottom")
```

As one can see, observed data is highly concentrated in Europe and China. Informations are extremely sparse everywhere else. This has obvious questions on how reliable the result will be if trained only on Europe and China and applied on regions with no coverage at all (such as Africa for instance).

Nonetheless, China offers a wide range of different environmental conditions, such as dry and cold regions in the north and central, temporal tropical regions in the south and wet temperate climate along the east coast. Even with a few data points a wide range of environmental conditions can be covered by China alone, which will be useful for spatial upscaling.

Nevertheless, desert regions are basically not represented, which thereby means that the results have to be viewed critically.

## 5.2 Spatial classification

To perform a spatial CV, we first need to group the observations. To do so we will use a kmeans unsupervised classifier using 5 centres (which returns 5 classes). To be able to process the data further, I will store the classes within the data frame.

> **Note**
>
> We need to specify the categorical format of the returned output using the `as.format()` function.

```{r geo kmeans}
clusters <- kmeans(
  dfs[, c(2, 3)], # Lon and lat as clustering parameters
  centers = 5 # 5 clusters
)

# Include class to df. Note: convert into factor format
dfs$class_spat <- as.factor(clusters$cluster)
```

To visualise the resulting clusters we can plot the classes after grouping the observations according to their geographical locations.

```{r plot geo clusters, fig.align='center', fig.cap="Figure 2: Clusters using a kmeans unsupervised learning model based on geographical locations of the observations form the dataset"}
ggplot() +
  
  #Plot coastline
  geom_sf(data = coast, colour = "black", size = 0.2) +

  # set extent in longitude and latitude
  # to draw map strictly bounded by the specified extent
  coord_sf(ylim = c(-60, 80), expand = FALSE) + 

  # plot points on map
  geom_point(data = dfs, aes(
    x = lon, 
    y = lat,
    colour = class_spat 
  ), size = 0.2) +
  
  # Adjust point size of legend
  guides(color = guide_legend(override.aes = list(size = 2))) +
  
  # Set colour
  scale_color_manual(name = "Class", values = brewer.pal(n = 5, "Set1")) +

  # Lable adjustments
  labs(
    x = "", y = "",
    title = "Spatial distribution of observations used in the dataset",
    subtitle = "Classification according to geographical location"
  ) +
  
  theme_bw() + 
  theme(legend.position = "right")
```

Following groups can be observed:

1.  Eastern Europe including western Asia, south eastern Sweden and one observation in Africa
2.  The American continent including the Iberian peninsula, the UK and the western part of France
3.  Central Europe including Denmark, south western Sweden and south Norway
4.  Northern Scandinavia
5.  China, India and Australia

In addition, we can show the distribution of leaf N over one class using a violin plot.

```{r violin spat, fig.align='center', fig.cap="Figure 3: Leaf nitrogen distribution of each class based on spatial classification using a violin plot"}
ggplot(data = dfs, aes(x = class_spat, y = leafN, fill = class_spat)) +
  geom_violin() + # Distribution using violin plot
  
  # Colour plot according to spatial distribution plot
  scale_fill_manual(values = brewer.pal(n = 5, "Set1")) +
  
  # Adding x & y axis label as well as title
  labs(
    x = "Class",
    y = "Leaf nitrogen",
    title = "Distribution of Leaf N over each class",
    subtitle = "Classification using geographical location"
  ) +
  
  theme_bw() +
  theme(legend.position = "") # Remove legend
```

From the distribution one can see that class 5 (China, India, Australia) offers the widest range of observations, where as class 1 (eastern Europe, western Asia, Africa) and 4 (northern Scandinavia) are heavily concentrated towards a small range of Leaf N values. class 2 (America, Iberian peninsula, UK) and 3 (central Europe) show very similar distribution with a two-peak distribution (a lower and a higher), where class 3 shows general higher leaf N values compared to 2.

## 5.3 Function

To extract the $R^2$ and $RMSE$ value from each CV-fold performed, I will first define a function which takes a data frame (the training set), the indices to consider for training and the indices to consider for validating as an input. The function will model the leaf N content given the predictors defined in chapter [4 Random CV] and validate the result by applying the generated model on the training set reduced to only contain data with the validation indices. The resulting predicted leaf N will be compared with the observed leaf N from the validation training set and the $R^2$ as well as the $RMSE$ value will be returned.

> **Note**
>
> Since the program returned an error when using `df[idx_val, 1]` as input for the `ranger::ranger()` function, I defined a training data set `df_train` and a testing data set `df_test` by taking a subset of the input data frame before using the `ranger::ranger()` function.

```{r cv function}
# create a function that trains a random forest model on a given set of rows and
# predicts on a disjunct set of rows
train_test_by_fold <- function(df, idx_train, idx_val){
  
  # Split into training and testing set
  df_train <- df[idx_train, ] # training set
  df_test <- df[idx_val, ] # testing set
  
  mod <- ranger::ranger(
    formula = df_train[, 1] ~ ., # target
    data = df_train[, c(4:9)], # predictors
  )

  pred <- predict(
    mod, # model
    data = df_test[ , c(4:9)] # corresponding rows + predictors
  )
  
  # Combine obs and pred into 1 df
  df_test$pred <- pred$predictions
  
  # Calculate RMSE
  rmse <- Metrics::rmse(
    actual = df_test$leafN,
    predicted = df_test$pred
  ) |> round(3)

  # Calculate R2
  rsq <- cor(df_test$leafN, df_test$pred, method = "pearson")^2 |> round(3)

  return(tibble(rsq = rsq, rmse = rmse))
}
```

## 5.4 Modelling

Given the classes derived in chapter x we can now split the data into a training and testing data for the 5-fold CV. We thereby define the test sets to contain only one class (e.g. class 1) and the training set to contain the remaining classes (e.g. 2 to 5). We will do that for every class available. This operation can be performed using the `purrr::map()` function.

```{r split geo data}
# Return Indices of every class except searched one
group_folds_train <- purrr::map(
  seq(length(unique(dfs$class_spat))), # Amount of classes
  ~ {
    dfs |>
      select(class_spat) |> # Select one class
      mutate(idx = 1:n()) |> # Generate index column
      filter(class_spat != .) |> # Filter for classes which is not currently selected
      pull(idx) # store row index of these classes in idx column 
  }
)

# Return indices of searched class
group_folds_test <- purrr::map(
  seq(length(unique(dfs$class_spat))), # Amount of classes
  ~ {
    dfs |>
      select(class_spat) |> # Select one class
      mutate(idx = 1:n()) |> # Generate index column
      filter(class_spat == .) |> # Filter for index of currently selected class
      pull(idx) # Store row index of theis class in idx column
  }
)
```

Given the training and testing set we can now apply the function created in chapter [5.3 Function] and the function `map2_dfr()` from the `purrr` package to combine $R^2$ and $RMSE$ value output from each CV-fold into one data frame.

```{r geo cv}
out <- purrr::map2_dfr(
  group_folds_train,
  group_folds_test, 
  ~ train_test_by_fold(df = dfs, .x, .y) # Function
) |>
  mutate(test_fold = 1:5) # CV fold row
```

To compare the result from the random 5-fold CV of chapter [4 Random CV] and the spatial 5-fold CV, we will combine the resulting output in one data frame.

```{r table spat}
stat <- stat |> mutate( # Add spatial cv metrics
  "Rsq_spat" = out$rsq,
  "RMSE_spat" = out$rmse
)

stat <- stat[c( # Reorder columns
  "Fold",
  "Rsq_ran",
  "Rsq_spat",
  "RMSE_ran",
  "RMSE_spat"
)] 

knitr::kable(
  stat,
  align = "lcccc",
  caption = "Table 3: Compareing R^2 and RMSE value from random and spatial 5-fold CV"
)
```

As one can see, the $R^2$ value form the random CV is way higher than the one from the spatial CV. In addition, the $RMSE$ from the random CV is lower than the one from the spatial CV. This means that the random CV out performs the spatial CV in general. I assume that the large differences occur due to the fact that the leaf N values are very specific inside their clusters. This means that a crucial specific condition is missing during the modelling operation as it was left out to perform a CV. As the random CV method takes every class from the spatial CV into account to split the data into 5 different folds, the chance is high that the model had a very diverse set of data points, including information from a region, which the spatial CV model had no access to. This could explain the better outcome of the random CV model.

# 6 Environmental CV

In a last step we will perform a CV based on environmental conditions.

## 6.1 Environmental classification

Like in chapter [5.2 Spatial classification] we will first group the observations based on a metric. However, this time we will use mean annual temperature and mean annual precipitation value to group the observations into 5 different classes. Again I will store the resulting prediction into the original data frame.

> **Note**
>
> We again need to specify the resulting values to be in a class format using the function `as.format()`.

```{r env kmeans}
clusters <- kmeans(
  dfs[, c(5, 6)], # Lon and lat as clustering parameters
  centers = 5 # 5 clusters
)

# Include class to df. Note: convert into factor format
dfs$class_env <- as.factor(clusters$cluster)
```

To visualise the resulting clusters we can plot the classes after grouping the observations according to their environmental conditions.

```{r plot env kmeans, fig.align='center', fig.cap="Figure 4: Clusters using a kmeans unsupervised learning model based on mean annual temperature and precipitation of the observations form the dataset"}
ggplot() +
  
  #Plot coastline
  geom_sf(data = coast, colour = "black", size = 0.2) +

  # set extent in longitude and latitude
  # to draw map strictly bounded by the specified extent
  coord_sf(ylim = c(-60, 80), expand = FALSE) + 

  # plot points on map
  geom_point(data = dfs, aes(
    x = lon, 
    y = lat,
    colour = class_env 
  ), size = 0.2) +
  
  # Adjust point size of legend
  guides(color = guide_legend(override.aes = list(size = 2))) +
  
  # Set colour
  scale_color_manual(name = "Class", values = brewer.pal(n = 5, "Set1")) +
  
  # Lable adjustments
  labs(
    x = "", y = "",
    title = "Spatial distribution of observations used in the dataset",
    subtitle = "Classification according to environmental conditions (T and Prec.)"
  ) +
  
  theme_bw() +
  theme(legend.position = "right")
```

To interpret the resulting classifications we need informations about the mean over one class of mean annual temperature and mean annual precipitation. As the training and testing set split will be done later on, I already generated a data frame containing the results.

```{r table, message=FALSE}
climate <- readr::read_csv("../data_raw/climate_table")

knitr::kable(
  climate,
  align = "lll",
  caption = "Table 4: Mean annual temperature and mean annual precipitation over all observations of a specific class",
  col.names = c("Class", "Average temperature [°C]", "Average precipitation [mm/yr]")
)
```

Following groups can be observed from the table and the plot above:

1.  Wet mid latitudes: Temperate regions with high precipitation
2.  Subtropics and boreal zones; dry regions (temperature less important)\
3.  Cold & wet mid latitudes: Cold temperate regions with high precipitation
4.  Warm & dry mid latitudes: Warm temperate regions with low precipitation
5.  Tropical; High temperature and precipitation

In addition, we can show the distribution of leaf N over one class using a violin plot.

```{r violin env, fig.align='center', fig.cap="Figure 5: Leaf nitrogen distribution of each class based on environmental classification using a violin plot"}
ggplot(data = dfs, aes(x = class_env, y = leafN, fill = class_env)) +
  geom_violin() + # Distribution using violin plot
  
  # Colour plot according to spatial distribution plot
  scale_fill_manual(values = brewer.pal(n = 5, "Set1")) +
  
  # Adding x & y axis label as well as title
  labs(
    x = "Class",
    y = "Leaf nitrogen",
    title = "Distribution of Leaf N over each class",
    subtitle = "Classification using environmental conditions (T and Prec.)"
  ) +
  
  theme_bw() +
  theme(legend.position = "") # Remove legend
```

Except class 2 (subtropics and boreal zones), which has to major maxima pretty close to each other, every class shows a major leaf N concentration with a tail towards higher leaf N concentrations. Class 1 (wet mid latitudes) shows a major and a minor concentration distribution, where as in class (cold and wet mid latitudes) show a very strong major leaf N concentration. Class 4 (warm and dry mid latitudes) and class 5 (tropical) show a pretty similar distribution.

## 6.2 Modelling

To perform the modelling step, we first need to split the data set into training and testing set, analogue to chapter [5.4 Modelling].

```{r split env data}
# Return Index of every class except searched one
group_folds_train_2 <- purrr::map(
  seq(length(unique(dfs$class_env))),
  ~ {
    dfs |>
      select(class_env) |>
      mutate(idx = 1:n()) |>
      filter(class_env != .) |>
      pull(idx)
  }
)

# Return index of searched one
group_folds_test_2 <- purrr::map(
  seq(length(unique(dfs$class_env))),
  ~ {
    dfs |>
      select(class_env) |>
      mutate(idx = 1:n()) |>
      filter(class_env == .) |>
      pull(idx)
  }
)
```

```{r climate table, echo=FALSE}
if(FALSE){
  # Generate climatological table
  class <- seq(1:5) # Classes
  climate <- data.frame(class)
  
  for(i in 1:5){
    data <- dfs[group_folds_test_2[[i]], ] # Generate df for class
    
    climate[i, "Temperatue"] <- round(mean(data$mat), 3) # Mean temperature
    climate[i, "Precipitation"] <- round(mean(data$map), 3) # Mean precipitation
  }
  
  # Save file
  write.csv(
    climate,
    file = "../data_raw/climate_table",
    row.names = FALSE
  )
}
```

Given the training and testing set we can now apply the function created in chapter [5.3 Function].

```{r env cv}
out_2 <- purrr::map2_dfr(
  group_folds_train_2,
  group_folds_test_2,
  ~ train_test_by_fold(df = dfs, .x, .y)
) |>
  mutate(test_fold = 1:5)
```

The resulting $R^2$ and $RMSE$ can now be incorporated into the table generated in chapter [5.4 Modelling] to make interpretations easier.

```{r table env}
stat <- stat |> mutate( # Add spatial cv metrics
  "Rsq_env" = out_2$rsq,
  "RMSE_env" = out_2$rmse
)

stat <- stat[c("Fold", # Reorder columns
  "Rsq_ran",
  "Rsq_spat",
  "Rsq_env",
  "RMSE_ran",
  "RMSE_spat",
  "RMSE_env"
)] 

knitr::kable(
  stat,
  align = "lcccccc",
  caption = "Figure 3: Compareing R^2 and RMSE value from random, spatial and environmental 5-fold CV"
)
```

As one can see, $R^2$ values of the environmental CV are in general higher than for a spatial CV, however still significantly lower than for a random CV. Differences in $RMSE$ values between environmental CV and spatial CV strongly depends on the fold. As for fold 1, spatial CV strongly outperforms environmental CV. The results on fold 2 show an opposite trend. $RMSE$ values for fold 3 show again a slightly better performance for spatial CV compared to environmental CV ans fold 4 shows again an opposite trend to fold 3. Fold 5 seams to return about the same $RMSE$ values. However, random CV returns a general lower, and thereby better, $RMSE$ value compared to both environmental CV and spatial CV.

Again I assume that the large differences between performance of random CV and environmental/spatial CV lays in the rather specific conditions of each class used for environmental/spatial CV. Since random CV tries to split the training data in N different by retaining the distribution across all predictors, random CV will be more generalisable as the model has seen a more diverse data set. As for the better performance of environmental CV compared to spatial CV one has to compare the leaf N distributions of environmentally based classification and spatially based classification. One can see that the distributions of environmentally based classification are more similar to each other compared to the distributions of spatially based classification. This means that the more similar distribution has the same effect as the random CV on being more generalisable and thereby less error susceptible to regions with highly specific climatic conditions.

# 7 Sources

Ludwig, Marvin, Alvaro Moreno-Martinez, Norbert Hölzel, Edzer Pebesma, and Hanna Meyer. 2023. “Assessing and Improving the Transferability of Current Global Spatial Prediction Models.” Global Ecology and Biogeography 32 (3): 356–68. <https://doi.org/10.1111/geb.13635>
